Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2025-01-16 ResponseLetter.tex

\documentclass[12pt]{extarticle}
\usepackage{geometry}
\geometry{
a4paper,
total={170mm,257mm},
left=20mm,
top=20mm,
headheight=12pt
}

\usepackage{xcolor}
\usepackage{amssymb,amsmath,amsthm}
\usepackage{natbib}
\usepackage{commath}
\usepackage{url}

\renewcommand{\d}{{\rm d}}
\newcommand{\e}{{\rm e}}

% Title page
\title{
	Response to reviewers: \\ Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs \\ 
}

@@

8. Are the tests for standing genetic variation essentially genetic drift? Variation is present and in a tumor cell population wildly fluctuating in size due to treatment you just hit on a "right" variant stochastically when tumor size is small (less cells) that is resistant and it's driven to fixation? If so, using genetic drift in this section is rooting language and another way of describing alternative scenarios/hypotheses for tumor rebound.

\textcolor{blue}{In our model, where aneuploidy is deleterious prior to drug treatment, standing variation is generated mis-segragation and purged by selection. Since aneuploids are carry a considerable fitness cost, random drift is not expected to effect standing genetic variation. Standing variation actually increases with the population size, whereas the effect of drift decreases with the population size. We now explain this in line XXX: "We assume that $c$ is not too small (Table 1), so that standing variation is generated by a balance between the generation of anuploids and selection against them, without the effect of genetic drift."
}

9. Do we know if it's reasonable to assume the division rates between cell types (i.e., "sensitive" vs. "mutant") are the same? Maybe ground that assumption into something we know, or just say we are not sure if that is reasonable or not. This relates to point 1 and rooting the reader into these parameter values with something real or being honest about the assumption they are equal.

\textcolor{blue}{ % change 40
We now ground this assumption in results from the literature (line XXX): "The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day. We take this to be the division rate for sensitive and assume that mutant cells, which are resistant to therapy, have the same division rate. Indeed, doubling times for tumors after relapse is lower than that of primary tumors for a variety of tissues \citep{tezuka2007growth,rodgers2024glioblastoma} and metastatic melanoma has been shown to grow faster then primary melonoma \citep{carlson2003tumor}."
}

@@

- methods: 'We estimate c=...' - It sounds like you will estimate c from the observed fraction of aneuploid cells but that is not what you describe below.

\textcolor{blue}{%change 5
We revised the text to make it clear that we estimate $c$ from an experimental result and $f$ from $c$ (line XXX): "To estimate fitness cost of aneuploidy, $c$, we note that Lukow et al. (2021) mixed sensitive and aneuploid A375 melanoma cells at $1:1$ ratio, cultured them in a drug-free environment, and observed the ratio evolve as a function of time with the aneuploid cells declining to $15\%$ after 24 days. Thus, our estimate for the fitness cost is $c=\abs{\log\left(0.15/(1-0.15)\right)/24}\approx0.07$ per day (Chevin 2011).
We estimate the fraction of aneuploid cancer cells in the pre-treatment environment using the formula $f=\tilde{u}\lambda_s / c$, which produces an estimate of $f=10^{-3}\times10^{-1}/0.07$, that is, $0.14\%$ of pre-treatment cancer cells have the beneficial aneuploidy of interest."
}

- methods: is there a justification for not including density-dependence for the sensitives as well? This would hasten their initial decline, which would compound with the reduced initial establishment probability of the mutants. It seems odd they can exert competition on the mutants but don't themselves feel any competition.

\textcolor{blue}{We have now added density-dependence for the sensitives. The results are unchanged.}

@@

- methods: I got a 404 error for the source code link: https://github.com/yoavram-lab/EvolutionaryRescue

\textcolor{blue}{
The repository will become available when the paper is accepted. Meanwhile, the source code for our analysis can be privately accessed here: \url{https://www.dropbox.com/scl/fi/hrvqdocsr67qur7k59wxw/StanaEtAlCode.zip?rlkey=n7bcr3usx5ym6q3s2qkotanhx}
}

@@

- results: I don't get the logic at the end of this paragraph, that your results are consistent with aneuploidy conferring resistance by decreasing the division rate. My understanding is that your result says that, holding aneuploid growth rate constant, increasing aneuploid division rate helps until you reach scenario 2 (effectively stationary/critical aneuploids) but beyond that doesn't help. I don't see how reducing division rate helps rescue.

\textcolor{red}{%change 52
In the scenario discribed decreasing the division rate also leads to a decrease in the death rate as the drugs affect the cancer cells during cellular division. We have expanded on the explanation of this scenario in the paper.
}

- results: if $N_a^*$ is similar to the detection threshold, then you predict that any tumour detected is very likely (eg, $>50\%$) to evolve resistance? Does this fit with data on resistance rates?

\textcolor{blue}{ % change 58
We believe that our model does fit with data on resistance rates, and the text now reads (line XXX): "Interestingly, the threshold between small and intermediate tumors, $N_a^*$, is similar to the tumor detection threshold of $4.19 \times 10^6$ cells for a wide variety of tumors~\citep{avanzini2019cancer}. We note that vemurafenib-treated melanomas (i.e. melanomas with sizes above the detection threshold) have a probability $>50\%$ to relapse \citep{piejko2023long,handa2022long}."
}

@@

- results: the final term of the recurrence time for small tumours is equivalent to the return time of Orr $\&$ Unckless 2014 with standing genetic variation (eqn 16), but different from their return time with new mutation (eqn 23). Why?

\textcolor{blue}{The return time with new mutation (eqn 23) from Orr $\&$ Unckless 2014 can be written as:
\begin{align*}
t_{return}&\sim\frac{1}{s-r}\log\left[\frac{2N_0s(s-r)}{r}\right]\\
&\sim \frac{1}{s-r}\log[2N_0(s-r)]+\frac{1}{s-r}\log\left(\frac{s}{r}\right)
\end{align*}
Additionally, when $|s-r|<r$ we have:
\begin{equation*}
\log s \approx \log r +\frac{s-r}{r}
\end{equation*}
which can be rewritten as:
\begin{align*}
\frac{1}{s-r}\log\left(\frac{s}{r}\right)\approx \frac{1}{r}
\end{align*}
As a result, the return time with new mutation can be written as:
\begin{equation*}
t_{return}\sim \frac{1}{r}+\frac{1}{s-r}\log[2N_0(s-r)]
\end{equation*}
which is the one-step equivalent to our recurrence time for small tumors.
}

- results: you mention that drugs increasing the death rate of sensitive cells delays recurrence, but given that mutations happen during division, it might be worth mentioning that drugs that decrease the division rate of sensitive cells by the same amount delay recurrence even more (holding mutant division rate constant - which is maybe why you didn't mention this for your parameterization, but seems like a worthwhile general point?).

\textcolor{blue}{% change 49
The text now reads (line XXX): "Drugs that increase the sensitive death rate, $\mu_s$, but do not affect the sensitive division rate, $\lambda_s$, delay cancer recurrence. Drugs that decrease the sensitive division rate delay cancer recurrence time even more, because they effectively decrease the mutation rate (assuming mutations occur during division)."
}

@@

- appendix C: I think it would be helpful to define $N_m^*$ and $N_a^*$ again in the appendix before using them

\textcolor{blue}{We have added the definitions of $N_m^*$ and $N_a^*$ in Appendix C. } %change 38

- appendix C: I would like to hear the justification for the initial expression of $\tau_a$ (integral of exponential) when $N>N_m^*$

\textcolor{blue}{%change 23
The justification, which is now included in Appendix C, is the following: 
\begin{align*}
\tau_a&=\int_0^\infty tf(t)dt=\int_0^\infty t\frac{d}{dt}F(t)dt=-\int_0^\infty t\frac{d}{dt}S(t)dt\\
&=-\left[tS(t)\right]_0^\infty+\int_0^\infty S(t)dt\\
&=\int_0^\infty S(t)dt,
\end{align*}
where $f(t)$, $F(t)$ and $S(t)$ are the probability density function, cumulative distribution function and survival function.
In our case, $S(t)=\exp\left(-\int_0^t R_a(z)dz\right)$.
Additionally, for $N\gg N_m^*$ we have $1-\e^{-Np_s}\approx1$. 
We have corrected a mistake, we now write $\tau_a=\int_0^\infty \e^{-\int_0^\tau R_a(t)\d t}\d\tau$
instead of $\tau_a=\int_0^\infty \e^{-R_a(\tau)}\d\tau$.
}

@@

- appendix C: similarly, I think it would be nice to hear some justification for the approximations in eqn C9, ie, can ignore direct rescue and rescue via aneuploidy very unlikely

\textcolor{blue}{
The justificaiont, also included now in Appendix C, is in the following. 
We know that
\begin{align*}
\tau_a=\int_0^\infty t\left(v\lambda_ap_ma_t+v\lambda_sp_ms_t\right)\frac{\exp\left[-\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_s t}-1}{\Delta_s}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}\right] }{1-\e^{-Np_s}}\d t.
\end{align*}
Additionally, if $N\ll N_a^*$ then the probability distribution of evolutionary rescue time is skewed toward small values (i.e. evolutionary rescue occurs conditioned on it happening). As a result, we have: 
\begin{align*}
\frac{\e^{\Delta_s t}-1}{\Delta_s}&=\frac{1+\Delta_s t+\mathcal{O}\left(\left(\Delta_st\right)^2\right)-1}{\Delta_s}=t+\mathcal{O}\left(\Delta_st^2\right)\\
\frac{\e^{\Delta_a t}-1}{\Delta_a}&=\frac{1+\Delta_a t+\mathcal{O}\left(\left(\Delta_at\right)^2\right)-1}{\Delta_a}=t+\mathcal{O}\left(\Delta_at^2\right)
\end{align*}
We observe that the first term in the exponential can be approximated to be zero and the second term as -$v\lambda_sNp_mt$ where $t\ll1$. As a result, we can approximate:
\begin{equation*}
\exp\left[-\frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_s t}-1}{\Delta_s}-\frac{\e^{\Delta_a t}-1}{\Delta_a}\right)-v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}\right] \sim 1
\end{equation*}
Additionally, since $N\ll N_a^*\ll N_m^*$, evolutionary rescue is more likely to occur through the trajectory $sensitive \rightarrow aneuploid \rightarrow mutant$. As a result,
\begin{equation*}
\tau_a\sim\int_0^\infty \frac{tv\lambda_ap_ma_t}{1-\e^{-Np_s}}\d t.
\end{equation*}
}

@@

\textcolor{blue}{In our model, we are interested in a specific type of aneuploidy which increases the fitness of cancer cells in an environment affected by anti-cancer drugs. As a consequence, we catalogued all other forms of aneuploids as sensitive.}

- Code Base
The python scripts used for simulation studies are provided in the supplementary package, however the github link is not accessible.
It would improve the reproducibility of the manuscript if the code to reproduce individual figures is provided.
Documenting the code as well as adding a short tutorial will improve the reusability of the scripts.

- In your model, mutations appear to only confer resistance, however, most mutations are deleterious. How easy is it to incorporate the effects of deleterious mutations in your model? This should be addressed in the discussion.

\textcolor{blue}{We consider the fitness of the sensitive, aneuploid or mutant cells to be the average of each population. As a result, mutations that are deleterious are taken into consideration through the mean fitness of each population. We have added this comment in the "Parametrisation" section change 50}

@@

- You report the exact parameters used in most figures. Sometimes it is difficult to follow why a specific value for a parameter was chosen. It would improve the readability if you included a table where the variation in the parameter used is noted and explained.

\textcolor{blue}{The variation in our parameters in noted in Table 1 and the new Table 2 and is obtained from the papers which are cited there.}

@@

Minor comments
Introduction:
* 10 millions people die: add globally

\textcolor{blue}{We have made the recommended change. }  % change 30

* mathematically equivalent: this is incorrect because fitness-valley crossing papers typically ignore population size dynamics, whereas evolutionary rescue is explicitly set in the context of a declining population size.

\textcolor{blue}{We removed the statement about the mathematical equivalence of valley crossing and evolutionary rescue.}

@@

* how multiple mutations contribute: several important papers on this subject are referenced later but not in this paragraph. Please revise, reviewing the existing theory and saying how the current theoretical work differs.

Methods:
* The methods only describe the simulations, and the results focus on analytical results that don't follow from the methods. I suggest expanding the "Methods" to discuss the general approach, model structure, and quantities derived (you can rename this as "Model") and then move to "Results".

\textcolor{blue}{
The Methods section has several subsections: "Evolutionary model" that describes the model, from which the analytic results are derived; "Stochastic simulations", which describes the simulations used to validate the analytic results; "Parameterization", which describes how we parameterize the model; and "Denstity-dependent growth", which described the extended model that includes density dependence. Note that the "Evolutionary model" subsection is not focused on the simulations but rather described the general model that we are studying.}

@@

* Delta would be better as "r" for the growth rate

\textcolor{blue}{The manuscript has over 300 mentions of $\Delta$, and we do not want to introduce any errors by changing it.}

@@

* The use of the terms "tolerant" (for negative growth) and "resistant" (for positive growth) is not standard (e.g., see different for fungi used by Berman $\&$ Krysan 2020 Nature Reviews Microbiology). Maybe alert the reader that "tolerant" is used in different ways in the literature and is used here solely for notational convenience to refer to mutations that, on their own, do not allow growth? Or replace with "sub-critical", a term already used in this context?

\textcolor{blue}{We have added a short explanation for "tolerant" and that it has other meanings in the literature. We also mention the equivalence to in "sub/super/critical" from Osmond et al 2020.} % change 47, 57

@@

* "antagonizing cell division" - The meaning of this isn't clear. Rephrase as "by extending the length of the cell cycle"?

\textcolor{blue}{We have made the recommended change from "antagonizing cell division" to "by extending the length of the cell cycle".}  % change 34

Results:
* Figure 7 - Add a numerical solution to equation (D3)? This should make it clearer that the non-monotonicity of the black curve is an artefact of the simplifying assumptions in (D4) and not a problem with the approach.

\textcolor{blue}{We have added the green line for eq. D3 to Figure 7 as suggested, it is very close to the blue line for eq. D4.}

* By the way, It wasn't clear to me what happened to the other large $\tau_a^r$ terms in the first portion of equation (D3) when approximating the equation into the form shown in (D4)? Could this be briefly clarified?

\textcolor{blue}{In the first term of (D3) we have: 
\begin{align*}
\frac{uv\lambda_a\lambda_s}{\Delta_s-\Delta_a}\left[\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}-\frac{\e^{\Delta_a\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_a-\Delta_m}\right]\approx\frac{uv\lambda_a\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_s\right)\left(\Delta_m-\Delta_a\right)}
\end{align*}
which is significantly smaller then the second term of (D3):
\begin{align*}
v\lambda_s\frac{\e^{\Delta_s\tau_a^r}-\e^{\Delta_m\tau_a^r}}{\Delta_s-\Delta_m}\approx\frac{v\lambda_s\e^{\Delta_m\tau_a^r}}{\left(\Delta_m-\Delta_a\right)}
\end{align*}
} % change 44

\textbf{Additional changes}

We have made the following additional changes:

$w_t\rightarrow s_t$. %change r1

$t\rightarrow \tau$. %change r2

The rates of the inhomogenous Poisson process are: %change r3
\begin{align*}
r_1\left(t\right)&=v\lambda_ap_ma_{t},\\ 
r_2\left(t\right)&=v\lambda_sp_ms_{t},
\end{align*}
and the expected number of successful mutants created by direct mutation and via aneuploidy until time $t$ is given by:
\begin{align*}
M_a\left(t\right)&=v\lambda_ap_m\int_0^ta_{z} d z = \frac{uv\lambda_s\lambda_aNp_m}{\Delta_s-\Delta_a}\left(\frac{\e^{\Delta_st}-1}{\Delta_s}-\frac{\e^{\Delta_at}-1}{\Delta_a}\right),\\ 
M_m\left(t\right)&=v\lambda_sp_m\int_0^ts_{z} d z = v\lambda_sNp_m\frac{\e^{\Delta_s t}-1}{\Delta_s}.
\end{align*}

Appendix C: all dummy variable used in integration are consistent. %change r4

Figure S6, S7, S8, S9: changed the subscripts of several variables from $w$ to $s$ . %change r5

"If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the rates at which the rescue mutations are generated can be written as" $\rightarrow$
"If a fraction $f$ of the cancer cells are aneuploid when the drug is administered then the expected number of successful mutants generated until time $t$ is given by". % change r6

### 2025-01-16 EvolutionaryRescue1.tex
continue editing the revision and response

\paragraph{Parameterization.}
We parametrize most of our simulations by considering melanoma cells and rely on \citet{rew2000cell} and \citet{bozic2013evolutionary} for the division and death rates, respectively. 
\citet{rew2000cell} report \textit{in vivo} measurements of the potential doubling times (the waiting time for the number of cells in the tumor to double, disregarding cell death) for a large set of cancer types. The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day. We select this to be the division rate for sensitive and mutant cells. This is a reasonable assumption given that we assume that the mutation makes the cancer cells immune to the chemotherapeutic drugs. %change 40

@@

\begin{figure}
\vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/ProliferationTime.pdf}
\caption{\textbf{Tumor size decreases the mean recurrence time.}
The mean time for the mutant cell population to reach size $N$, the initial number of cancer cells.
Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with 95\% confidence interval obtained with bootstrapping, see Appendix G) for intermediary $N$. The simulations converge to \cref{eq:t2det} (blue dashed line) for large values of $N$. The green line represents the numerical solution of \cref{eq:t2det2024}. % change 44  
Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
\label{proliferationFigure}
\end{figure}

>>>>

\paragraph{Parameterization.}
We parametrize most of our simulations by considering melanoma cells and rely on \citet{rew2000cell} and \citet{bozic2013evolutionary} for the division and death rates, respectively. 
\citet{rew2000cell} report \textit{in vivo} measurements of the potential doubling times (the waiting time for the number of cells in the tumor to double, disregarding cell death) for a large set of cancer types. The division rate is obtained as $\lambda=\log{2} / T \approx 0.1$ per day. We take this to be the division rate for sensitive and assume that mutant cells, which are resistant to therapy, have the same division rate. Indeed, doubling times for tumors after relapse is lower than that of primary tumors for a variety of tissues \citep{tezuka2007growth,rodgers2024glioblastoma} and metastatic melanoma has been shown to grow faster then primary melonoma \citep{carlson2003tumor}. %change 40

@@

\begin{figure}
\vspace*{1\baselineskip}
\includegraphics[width=1\textwidth]{Figures/ProliferationTime.pdf}
\caption{\textbf{Tumor size decreases the mean recurrence time.}
The mean time for the mutant cell population to reach size $N$, the initial number of cancer cells.
Our inhomogeneous Poisson-process approximation (solid black line, \cref{meanproliferationtime}) is in agreement with simulation results (red markers with error bars for 95\% CI) for intermediate size $N$.
Simulation results converge to \cref{eq:t2det} (blue dashed line) for large values of $N$. The green line represents the numerical solution of \cref{eq:t2det2024}. % change 44  
Parameters: $\lambda_s=0.1,\lambda_a=0.0899,\lambda_m=0.1,\mu_s=0.14,\mu_a=0.09,\mu_m=0.09, u=10^{-2}, v=10^{-7}$.}
\label{proliferationFigure}
\end{figure}
